Systemix Article: Combined MAG-SMAG Profiling

### 2024-06-03 R/paper.Rmd
updated to manuscript version 1.2

#count number of MAGs per ecosystem
mags_depth_spec_eco_n <- mags_metadata_bac %>%
  group_by(ecosystem) %>%
  dplyr::summarise(mags = n_distinct(mag_name))
mags_depth_spec_eco <- left_join(mags_depth_spec_eco, mags_depth_spec_eco_n, by = c("ecosystem"))
rm(mags_depth_spec_eco_n)

@@

### --EXPLORE OTHER TAXA-- ###
#inspect unannotated family-level bins (uFGBs)? #1166
#what about novel bacteria at genus level?
mags_depth_novel_gen <- mags_metadata_bac %>%
  dplyr::filter(genus == "") %>%
  dplyr::group_by(phylum, genus, ecosystem) %>%
  dplyr::summarise(gen_unk = n()) %>%
  #dplyr::arrange(desc(gen)) %>%
  ungroup() %>%
  dplyr::group_by(genus) %>%
  dplyr::mutate(tot_gen = sum(gen_unk)) %>%
  dplyr::mutate(perc = gen_unk * 100 / tot_gen)

#get unique classes of unknown genus by ecosystem
#Leptospirillia, Sulfobacillia
mags_md_unk_gen2 <- mags_metadata_bac %>%
  dplyr::filter(genus == "") %>%
  dplyr::group_by(class, ecosystem) %>%
  tally() %>%
  #slice(n()) %>%
  ungroup()
unique(mags_md_unk_gen2[mags_md_unk_gen2[, "ecosystem"] == "Biomining", "class"])

#unique genus from biomining: Acidithrix, Ferrithrix, SCQM01
unique(mag_samples[mag_samples[, "family"] != "", "genus"])
mags_metadata_bac %>% filter(grepl("Acidithrix", genus, ignore.case = TRUE) == TRUE)

plot_md_unk_cla <- ggplot(mags_md_unk_gen2[mags_md_unk_gen2$n > 10, ],
                          aes(x = reorder(class, n), y = n, fill = ecosystem)) +
  geom_bar(stat = "identity") +
  labs(x = "Class", y = "Count") +
  scale_y_continuous(limits = c(0, 250)) +
  theme_minimal() +
  coord_flip() +
  facet_wrap(~ecosystem)
ggsave("plot_md_unk_class_eco.pdf", plot_md_unk_cla, height = 7, width = 10)

#class of interest for acidic soils:
#Alphaproteobacteria, Gammaproteobacteria
#Acidobacteriae, Thermoleophilia, Sulfobacillia, Acidimicrobiia
table(mags_metadata_bac[mags_metadata_bac[, "phylum"] == "Proteobacteria", "class"])
table(mags_metadata_bac[mags_metadata_bac[, "class"] == "Acidimicrobiia", "order"])

#family of interest reported for Systemix [ref 2022]:
#Pseudomonadaceae (mayor), Flavobacteriaceae, Erwiniaceae
mags_md_fams_acid <- mags_metadata_bac %>%
  dplyr::filter(family == "Pseudomonadaceae") %>%
  dplyr::group_by(ecosystem) %>%
  dplyr::summarise(count = n())
```

@@

```{r magbiomqc}
# Read QC file from CheckM
mags_qc <- read.csv("2-statistics/checkm_parsed_bins.tsv", sep = "\t")

# Select current sample
mags_qc <- mags_qc %>% filter(grepl("Refined-S15", bin))

# Parse QCs
mags_qc_hq <- mags_qc %>%
  dplyr::filter(completeness >= 90 & contamination <= 5) %>%
  dplyr::summarise(mags = n(), contigs = sum(contigs))
mags_qc_hq$label = "High Quality"

mags_qc_mq <- mags_qc %>%
  dplyr::filter(completeness < 90 & completeness >= 50 & contamination <= 10) %>%
  dplyr::summarise(mags = n(), contigs = sum(contigs))
mags_qc_mq$label = "Medium Quality"

mags_qc_lq <- mags_qc %>%
  dplyr::filter(completeness < 50 & contamination <= 10) %>%
  dplyr::summarise(mags = n(), contigs = sum(contigs))
mags_qc_lq$label = "Low Quality"

mags_qcs <- rbind(mags_qc_hq, mags_qc_mq, mags_qc_lq)
rm(mags_qc_hq, mags_qc_mq, mags_qc_lq)

# Plot
plot_mags_qc <- ggplot(mags_qcs, aes(x = mags, y = fct_reorder(label, mags), fill = label)) +
  geom_col(position = "dodge2") +
  scale_fill_manual(values = c("#DB7971", "#5CA650", "#859AC7")) +
  labs(x = "Number of MAGs", y = "") +
  theme_classic() +
  theme(legend.position = "none")
ggsave("plot_biom_MAGs_QC.pdf", plot_mags_qc, height = 3, width = 5)
```

@@

# Export tree to file
if (file.exists("1-taxonomy/trees-subset/gtdbtk.DASTool-S15.tree") == FALSE) {
  ape::write.tree(tree_s15, "1-taxonomy/trees-subset/gtdbtk.DASTool-S15.tree")
}
rm(tree)

@@

# Save attributes in a dataframe
tree_qc_df <- tibble(Bin = tree_tips, Completeness = tree_com,
                     Contamination = tree_con, Strain_Het = tree_st,
                     Genome_Size = tree_gs, GC = tree_gc, Pred_Genes = tree_pg)
rm(tree_tips, tree_com, tree_con, tree_st, tree_gs, tree_gc, tree_pg)

@@

fig_1 <- (fig_1a / fig_1b) +
  plot_annotation(tag_levels = list(c("a", "e", "b", "c", "d", "", "f"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_1.pdf", fig_1, height = 14, width = 14)
#ggsave("fig_1.png", fig_1, height = 14, width = 14)
rm(fig_1a, fig_1b)

#sup 1
fig_s1 <- { plot_depth_top_phyla |
    (plot_depth_usgbs / plot_depth_spec + plot_layout(heights = c(0.85, 0.15))) } +
  plot_layout(ncol = 2, widths = c(0.25, 1)) +
  plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag =  element_text(face = "bold"))
fig_s1[[2]][[1]] <- fig_s1[[2]][[1]] + theme(axis.title.y = element_text(vjust = 0))
#fig_s1[[2]][[2]] <- fig_s1[[2]][[2]] + theme(axis.text.y = element_text(vjust = 0))
##use wrap_plots() to correct axis line/labels ??
ggsave("mags-all/fig_s1.pdf", fig_s1, height = 8, width = 16)
#ggsave("mags-all/fig_s1.png", fig_s1, height = 8, width = 16)
```

@@

# Inspect refdb
#load("mags-all/5-SNPs/dndscv/buildref/S15_refcds.rda")
#print(RefCDS[[1]])
#rm(gr_genes, RefCDS)

@@

# Run dndscv (runtime ~8 min, high memory usage) and save object
#snp_dndsout <- dndscv(snp_mutations, refdb = "mags-all/5-SNPs/dndscv/buildref/S15_refcds.rda",
#                      max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf,
#                      cv = NULL, outmats = TRUE)
#save(snp_dndsout, file = "mags-all/5-SNPs/dndscv/run/S15_dndsout.rda")
#save(substmodel, file = "mags-all/5-SNPs/dndscv/run/S15_substmodel.rda")
#load("mags-all/5-SNPs/dndscv/run/S15_dndsout.rda")
#load("mags-all/5-SNPs/dndscv/run/S15_substmodel.rda")
rm(snp_mutations)

@@

# Save matrix with dN/dS: driver discovery or genes under positive selection (PS)
snp_selcv <- snp_dndsout$sel_cv
rm(snp_dndsout)

@@

snp_taxabins <- tibble(phylum = character(), genus = character(), species = character(),
                       completeness = numeric(), contamination = numeric(), predicted_genes = numeric())
for (i in 1:nrow(snp_dnds)) {
  #print(sprintf("%i -> bin: %s", i, fx_dndsTaxaBin(snp_dnds$bin[i])))
  snp_taxabins <- rbind(snp_taxabins, fx_dndsTaxaBin(snp_dnds$bin[i]))
}
snp_dnds['phylum'] <- snp_taxabins$phylum
snp_dnds['genus'] <- snp_taxabins$genus
snp_dnds['species'] <- snp_taxabins$species
snp_dnds['completeness'] <- snp_taxabins$completeness
snp_dnds['contamination'] <- snp_taxabins$contamination
snp_dnds['predicted_genes'] <- snp_taxabins$predicted_genes
rm(snp_taxabins)

@@

# Subset PS genes to plot mutations -> synonymous/missense (non-syn.)/nonsense
#discard HPs (by name)
#genes with at least 4 SNPs
#mark genes reaching significance (q < 0.05)
#mark genes: KJCAEPPB_04437 -> nickel pathway; NEBJFEFK_01910 -> sulfur pathway
snp_psgenes_sub <- snp_psgenes %>%
  dplyr::filter(nchar(name) < 14) %>%
  dplyr::mutate(snps = n_syn + n_mis + n_non) %>%
  dplyr::filter(snps >= 4) %>%
  dplyr::mutate(signif = ifelse(pmis_cv < 0.05, "*", "")) %>%
  dplyr::mutate(path = case_when(gene_name == "KJCAEPPB_04437" ~ "Ni",
                                 gene_name == "NEBJFEFK_01910" ~ "S", TRUE ~ "")) %>%
  distinct(name, .keep_all = TRUE) %>% #remove repeated names
  pivot_longer(cols = c(n_syn, n_mis, n_non), names_to = "mut", values_to = "n_mut")

@@

#add completeness & contamination barplots
snp_plot_psbins2 <- ggplot(snp_dnds, aes(x = completeness, y = reorder(bin, num), fill = bin)) +
  geom_col(fill = "#3c4d76") +
  geom_vline(xintercept = 100, linetype = 1, linewidth = 0.2) +
  labs(x = "Completeness", y = "") +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())
snp_plot_psbins3 <- ggplot(snp_dnds, aes(x = contamination, y = reorder(bin, num), fill = bin)) +
  geom_col(fill = "#763c4d") +
  geom_vline(xintercept = max(snp_dnds$contamination), linetype = 1, linewidth = 0.2) +
  labs(x = "Contamination", y = "") +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())

#merge
snp_plot_psbins <- wrap_plots(snp_plot_psbins, snp_plot_psbins2, snp_plot_psbins3,
                              nrow = 1, ncol = 3, widths = c(1, 0.15, 0.15))
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psbins.pdf", snp_plot_psbins, height = 9, width = 12)
rm(snp_plot_psbins2, snp_plot_psbins3)

@@

# Plot PS genes mutations per type
snp_plot_muts <- ggplot(snp_psgenes_sub, aes(x = n_mut, y = reorder(name, wmis_cv), fill = mut)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(name = "", labels = c("mis", "non", "syn"),
                    values = c("#5db4b2", "#b687b6", "#2374a4")) +
  labs(x = "Number of mutations", y = "") +
  theme_classic() +
  theme(legend.position = c(0.85, 0.15),
        strip.placement = "outside",
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psmuts.pdf", snp_plot_muts, height = 11, width = 7)

#merge
snp_plot_psgenes <- (snp_plot_psgenes | snp_plot_muts) +
  plot_layout(nrow = 1, ncol = 2, widths = c(0.55, 0.5))
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psgenmuts.pdf", snp_plot_psgenes, height = 11, width = 11)
rm(snp_plot_muts)

@@

# Infer pathway from product and add dndscv attributes
snp_nspath <- tibble(gene_name = character(), pathway = character())
snp_nsstat <- tibble(wmis_cv = numeric(), pmis_cv = numeric())
for (i in 1:nrow(snp_nsgenes)) {
  #use dataframes, not lists
  snp_nspath <- snp_nspath %>%
    add_row(gene_name = snp_nsgenes$gene_name[i], pathway = fx_dndsNsPathway(snp_nsgenes$product[i]))
	snp_nsstat <- rbind(snp_nsstat, fx_dndsNsStats(snp_nsgenes$gene_name[i]))
}
snp_nsgenes <- left_join(snp_nsgenes, snp_nspath, by = c("gene_name"))
snp_nsgenes <- left_join(snp_nsgenes, snp_nsstat, by = c("gene_name"))
rm(snp_nspath, snp_nsstat)

# Filter NS genes not called in dndscv
#omit significance (pmis_cv < 0.05)
#discard nitrogen (not enough genes)
#change %2C to comma
snp_nsgenes <- snp_nsgenes %>%
  dplyr::filter(!(is.na(wmis_cv)) & wmis_cv < 1 & wmis_cv != 0 & pathway != "Nitrogen") %>%
  dplyr::mutate(product = str_replace_all(product, "%2C", ","))

@@

# Merge NS genes to unique products
snp_nsgenes_sub <- snp_nsgenes %>%
  dplyr::group_by(product, pathway) %>%
  dplyr::summarise(dnds = mean(wmis_cv))
snp_nsgenes_sub_col <- c("#8c2421", "#666666", "#4f4fad", "#ddad3b")

# Lollipop plot of NS genes dN/dS x-axis range 0-2, y-axis PS genes vs selected CDS
snp_plot_nsgensub <- ggdotchart(snp_nsgenes_sub, x = "product", y = "dnds", color = "pathway",
                              xlab = "NS gene product", ylab = "Mean dN/dS", y.text.col = FALSE,
                              rotate = TRUE, palette = snp_nsgenes_sub_col,
                              sorting = "descending", ggtheme = theme_pubr()) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2)
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_nsgenesprod.pdf", snp_plot_nsgensub, height = 9, width = 10)

# Global gene estimation of dn/ds (<1 & >1) per number of genes and phylum
snp_estim <- snp_selcv %>%
  dplyr::select(gene_name, wmis_cv, pmis_cv) %>%
  dplyr::filter(wmis_cv != 0)

@@

# Add taxa & QC
snp_estim <- left_join(snp_estim, snp_dnds, by = c("bin")) %>%
  dplyr::select(-c("contig", "num", "genus", "species"))

# Normalize dN/dS per gene by MAG contamination, then group by phylum
snp_estim_phy <- snp_estim %>%
  dplyr::mutate(dnds_adj = wmis_cv * (1 - contamination / 100)) %>%
  dplyr::group_by(phylum, bin) %>%
  dplyr::summarise(snps = n(), dnds_sum = sum(dnds_adj)) %>%
  ungroup() %>%
  dplyr::mutate(dnds_glob = dnds_sum / snps) %>% #divide by total genes and top 3 phyla
  dplyr::filter(phylum == "Proteobacteria" | phylum == "Acidobacteriota" | phylum == "Actinobacteriota")

# Sort phylum by median and set color
snp_estim_sort_phy <- snp_estim_phy %>%
  dplyr::group_by(phylum) %>%
  dplyr::summarise(n = median(dnds_glob)) %>%
  dplyr::arrange(desc(n))
snp_estim_sort_phy$color <- rep("#4c6194", 3)

@@

# Load lists of PS and NS genes (with KEGG KOs from KOfamScan)
snp_kops <- read.csv("mags-all/5-SNPs/dndscv/results/kofam/ps_genes_kos.txt", sep = "\t", header = FALSE, stringsAsFactors = FALSE)
snp_kons <- read.csv("mags-all/5-SNPs/dndscv/results/kofam/ns_genes_kos.txt", sep = "\t", header = FALSE, stringsAsFactors = FALSE)

# Reduce to unique KOs, saving scores
snp_kops <- snp_kops %>%
  dplyr::group_by(V3, V7) %>%
  dplyr::summarise(sum_ko = n(), score_avg = mean(V4), .groups = "drop") %>%
  dplyr::rename(KO = V3, KO_definition = V7) %>%
  dplyr::mutate(KO_def = paste0("(", KO, ") - ", KO_definition)) %>%
  dplyr::mutate(dNdS = "PS")
snp_kons <- snp_kons %>%
  dplyr::group_by(V3, V7) %>%
  dplyr::summarise(sum_ko = n(), score_avg = mean(V4), .groups = "drop") %>%
  dplyr::rename(KO = V3, KO_definition = V7) %>%
  dplyr::mutate(KO_def = paste0("(", KO, ") - ", KO_definition)) %>%
  dplyr::mutate(dNdS = "NS")
#merge dataframe and filter
snp_kos <- bind_rows(snp_kops, snp_kons)
snp_kos <- snp_kos %>%
  dplyr::filter(sum_ko > 10)
rm(snp_kops, snp_kons) #clean memory

@@

# Filter genome bin with PS selection
snp_cirmag <- snp_estim %>% dplyr::filter(grepl("S15.49", bin))

# Set genomic attributes: gc, contigs
snp_cirmagname <- "SPAdes-MetaBAT2Refined-S15.49"
snp_cirsize <- c(0, 2703949)
snp_cirran <- c(0, 0.5, 1, 1.5, 2, 2.5, 2.7) * 10^6

>>>>

#count number of MAGs per ecosystem
mags_depth_spec_eco_n <- mags_metadata_bac %>%
  group_by(ecosystem) %>%
  dplyr::summarise(mags = n_distinct(mag_name))
mags_depth_spec_eco <- left_join(mags_depth_spec_eco, mags_depth_spec_eco_n, by = c("ecosystem"))
rm(mags_depth_spec_eco_n) #clean memory

@@

### --EXPLORE OTHER TAXA-- ###
##inspect unannotated family-level bins (uFGBs)? #1166
##what about novel bacteria at genus level?
#mags_depth_novel_gen <- mags_metadata_bac %>%
#  dplyr::filter(genus == "") %>%
#  dplyr::group_by(phylum, genus, ecosystem) %>%
#  dplyr::summarise(gen_unk = n()) %>%
#  #dplyr::arrange(desc(gen)) %>%
#  ungroup() %>%
#  dplyr::group_by(genus) %>%
#  dplyr::mutate(tot_gen = sum(gen_unk)) %>%
#  dplyr::mutate(perc = gen_unk * 100 / tot_gen)
#
##get unique classes of unknown genus by ecosystem
##Leptospirillia, Sulfobacillia
#mags_md_unk_gen2 <- mags_metadata_bac %>%
#  dplyr::filter(genus == "") %>%
#  dplyr::group_by(class, ecosystem) %>%
#  tally() %>%
#  #slice(n()) %>%
#  ungroup()
#unique(mags_md_unk_gen2[mags_md_unk_gen2[, "ecosystem"] == "Biomining", "class"])
#
##unique genus from biomining: Acidithrix, Ferrithrix, SCQM01
#unique(mag_samples[mag_samples[, "family"] != "", "genus"])
#mags_metadata_bac %>% filter(grepl("Acidithrix", genus, ignore.case = TRUE) == TRUE)
#
#plot_md_unk_cla <- ggplot(mags_md_unk_gen2[mags_md_unk_gen2$n > 10, ],
#                          aes(x = reorder(class, n), y = n, fill = ecosystem)) +
#  geom_bar(stat = "identity") +
#  labs(x = "Class", y = "Count") +
#  scale_y_continuous(limits = c(0, 250)) +
#  theme_minimal() +
#  coord_flip() +
#  facet_wrap(~ecosystem)
#ggsave("plot_md_unk_class_eco.pdf", plot_md_unk_cla, height = 7, width = 10)
#
##class of interest for acidic soils:
##Alphaproteobacteria, Gammaproteobacteria
##Acidobacteriae, Thermoleophilia, Sulfobacillia, Acidimicrobiia
#table(mags_metadata_bac[mags_metadata_bac[, "phylum"] == "Proteobacteria", "class"])
#table(mags_metadata_bac[mags_metadata_bac[, "class"] == "Acidimicrobiia", "order"])
#
##family of interest reported for Systemix [ref 2022]:
##Pseudomonadaceae (mayor), Flavobacteriaceae, Erwiniaceae
#mags_md_fams_acid <- mags_metadata_bac %>%
#  dplyr::filter(family == "Pseudomonadaceae") %>%
#  dplyr::group_by(ecosystem) %>%
#  dplyr::summarise(count = n())
```

@@

```{r magbiomqc}
# Read QC file from CheckM
mags_qc <- read.csv("mags-all/2-statistics/checkm_parsed_bins.tsv", sep = "\t")

# Select S15 sample
mags_qc <- mags_qc %>% filter(grepl("Refined-S15", bin))

# Parse QCs
mags_qc_hq <- mags_qc %>%
  dplyr::filter(completeness >= 90 & contamination <= 5) %>%
  dplyr::summarise(mags = n(), contigs = sum(contigs),
                   avg_comp = mean(completeness), avg_cont = mean(contamination))
mags_qc_hq$label = "High Quality"
mags_qc_mq <- mags_qc %>%
  dplyr::filter(completeness < 90 & completeness >= 50 & contamination <= 10) %>%
  dplyr::summarise(mags = n(), contigs = sum(contigs),
                   avg_comp = mean(completeness), avg_cont = mean(contamination))
mags_qc_mq$label = "Medium Quality"
mags_qc_lq <- mags_qc %>%
  dplyr::filter(completeness < 50 & contamination <= 10) %>%
  dplyr::summarise(mags = n(), contigs = sum(contigs),
                   avg_comp = mean(completeness), avg_cont = mean(contamination))
mags_qc_lq$label = "Low Quality"
#merge
mags_qcs <- rbind(mags_qc_hq, mags_qc_mq, mags_qc_lq)
rm(mags_qc_hq, mags_qc_mq, mags_qc_lq) #clean memory

# Plot
plot_mags_qc <- ggplot(mags_qcs, aes(x = mags, y = fct_reorder(label, mags), fill = label)) +
  geom_col(position = "dodge2") +
  scale_fill_manual(values = c("#DB7971", "#5CA650", "#859AC7")) +
  labs(x = "Number of MAGs", y = "") +
  theme_classic() +
  theme(legend.position = "none")
ggsave("mags-all/2-statistics/plots/plot_biom_MAGs_QC.pdf", plot_mags_qc, height = 3, width = 5)
```

@@

# Export tree to file
if (file.exists("1-taxonomy/trees-subset/gtdbtk.DASTool-S15.tree") == FALSE) {
  ape::write.tree(tree_s15, "1-taxonomy/trees-subset/gtdbtk.DASTool-S15.tree")
}
rm(tree) #clean memory

@@

# Save attributes in a dataframe
tree_qc_df <- tibble(Bin = tree_tips, Completeness = tree_com,
                     Contamination = tree_con, Strain_Het = tree_st,
                     Genome_Size = tree_gs, GC = tree_gc, Pred_Genes = tree_pg)
rm(tree_tips, tree_com, tree_con, tree_st, tree_gs, tree_gc, tree_pg) #clean memory

@@

fig_1 <- (fig_1a / fig_1b) +
  plot_annotation(tag_levels = list(c("a", "e", "b", "c", "d", "", "f"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_1.pdf", fig_1, height = 14, width = 14)
#ggsave("fig_1.png", fig_1, height = 14, width = 14)
rm(fig_1a, fig_1b) #clean memory

#sup 1
fig_s1 <- { plot_depth_top_phyla |
    (plot_depth_usgbs / plot_depth_spec + plot_layout(heights = c(0.85, 0.15))) } +
  plot_layout(ncol = 2, widths = c(0.25, 1)) +
  plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag =  element_text(face = "bold"))
fig_s1[[2]][[1]] <- fig_s1[[2]][[1]] + theme(axis.title.y = element_text(vjust = 0))
#fig_s1[[2]][[2]] <- fig_s1[[2]][[2]] + theme(axis.text.y = element_text(vjust = 0))
#fig_s1 <- wrap_plots(plot_depth_usgbs, plot_depth_spec, nrow = 2, ncol = 1, heights = c(0.85, 0.15))
ggsave("mags-all/fig_s1.pdf", fig_s1, height = 9, width = 18)
#ggsave("mags-all/fig_s1.png", fig_s1, height = 9, width = 18)
```

@@

# Inspect refdb
#load("mags-all/5-SNPs/dndscv/buildref/S15_refcds.rda")
#print(RefCDS[[1]])
#rm(gr_genes, RefCDS) #clean memory

@@

# Run dndscv (runtime ~8 min, high memory usage) and save object
#snp_dndsout <- dndscv(snp_mutations, refdb = "mags-all/5-SNPs/dndscv/buildref/S15_refcds.rda",
#                      max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf,
#                      cv = NULL, outmats = TRUE)
#save(snp_dndsout, file = "mags-all/5-SNPs/dndscv/run/S15_dndsout.rda")
#save(substmodel, file = "mags-all/5-SNPs/dndscv/run/S15_substmodel.rda")
#load("mags-all/5-SNPs/dndscv/run/S15_dndsout.rda")
#load("mags-all/5-SNPs/dndscv/run/S15_substmodel.rda")
rm(snp_mutations) #clean memory

@@

# Save matrix with dN/dS: driver discovery or genes under positive selection (PS)
snp_selcv <- snp_dndsout$sel_cv
rm(snp_dndsout) #clean memory

@@

snp_taxabins <- tibble(phylum = character(), genus = character(), species = character(),
                       completeness = numeric(), contamination = numeric(), predicted_genes = numeric())
for (i in 1:nrow(snp_dnds)) {
  #print(sprintf("%i -> bin: %s", i, fx_dndsTaxaBin(snp_dnds$bin[i])))
  snp_taxabins <- rbind(snp_taxabins, fx_dndsTaxaBin(snp_dnds$bin[i]))
}
snp_dnds['phylum'] <- snp_taxabins$phylum
snp_dnds['genus'] <- snp_taxabins$genus
snp_dnds['species'] <- snp_taxabins$species
snp_dnds['completeness'] <- snp_taxabins$completeness
snp_dnds['contamination'] <- snp_taxabins$contamination
snp_dnds['predicted_genes'] <- snp_taxabins$predicted_genes
rm(snp_taxabins) #clean memory

@@

# Subset PS genes to plot mutations -> synonymous/missense (non-syn.)/nonsense
#discard hypothetical proteins (HPs: according to name)
#genes with at least 4 SNPs
#mark genes reaching significance (q < 0.05)
#mark genes: KJCAEPPB_04437 -> nickel pathway; NEBJFEFK_01910 -> sulfur pathway
snp_psgenes_sub <- snp_psgenes %>%
  dplyr::filter(nchar(name) < 14) %>%
  dplyr::mutate(snps = n_syn + n_mis + n_non) %>%
  dplyr::filter(snps >= 4) %>%
  dplyr::mutate(signif = ifelse(pmis_cv < 0.05, "*", "")) %>%
  dplyr::mutate(path = case_when(gene_name == "KJCAEPPB_04437" ~ "Ni",
                                 gene_name == "NEBJFEFK_01910" ~ "S", TRUE ~ "")) %>%
  distinct(name, .keep_all = TRUE) %>% #remove repeated names
  pivot_longer(cols = c(n_syn, n_mis, n_non), names_to = "mut", values_to = "n_mut")

@@

# Add completeness & contamination barplots, then merge
snp_plot_psbins2 <- ggplot(snp_dnds, aes(x = completeness, y = reorder(bin, num), fill = bin)) +
  geom_col(fill = "#3c4d76") +
  geom_vline(xintercept = 100, linetype = 1, linewidth = 0.2) +
  labs(x = "Completeness", y = "") +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())
snp_plot_psbins3 <- ggplot(snp_dnds, aes(x = contamination, y = reorder(bin, num), fill = bin)) +
  geom_col(fill = "#763c4d") +
  geom_vline(xintercept = max(snp_dnds$contamination), linetype = 1, linewidth = 0.2) +
  labs(x = "Contamination", y = "") +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())

snp_plot_psbins <- wrap_plots(snp_plot_psbins, snp_plot_psbins2, snp_plot_psbins3,
                              nrow = 1, ncol = 3, widths = c(1, 0.15, 0.15))
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psbins.pdf", snp_plot_psbins, height = 9, width = 12)
rm(snp_plot_psbins2, snp_plot_psbins3) #clean memory

@@

# Plot PS genes mutations per type, then merge
snp_plot_muts <- ggplot(snp_psgenes_sub, aes(x = n_mut, y = reorder(name, wmis_cv), fill = mut)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(name = "", labels = c("mis", "non", "syn"),
                    values = c("#5db4b2", "#b687b6", "#2374a4")) +
  labs(x = "Number of mutations", y = "") +
  theme_classic() +
  theme(legend.position = c(0.85, 0.15),
        strip.placement = "outside",
        axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank())
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psmuts.pdf", snp_plot_muts, height = 11, width = 7)

snp_plot_psgenes <- (snp_plot_psgenes | snp_plot_muts) +
  plot_layout(nrow = 1, ncol = 2, widths = c(0.55, 0.5))
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_psgenmuts.pdf", snp_plot_psgenes, height = 11, width = 11)
rm(snp_plot_muts) #clean memory

@@

# Infer pathway from product and add dndscv attributes
snp_nspath <- tibble(gene_name = character(), pathway = character())
snp_nsstat <- tibble(wmis_cv = numeric(), pmis_cv = numeric())
for (i in 1:nrow(snp_nsgenes)) {
  #use dataframes, not lists
  snp_nspath <- snp_nspath %>%
    add_row(gene_name = snp_nsgenes$gene_name[i], pathway = fx_dndsNsPathway(snp_nsgenes$product[i]))
	snp_nsstat <- rbind(snp_nsstat, fx_dndsNsStats(snp_nsgenes$gene_name[i]))
}
snp_nsgenes <- left_join(snp_nsgenes, snp_nspath, by = c("gene_name"))
snp_nsgenes <- left_join(snp_nsgenes, snp_nsstat, by = c("gene_name"))
rm(snp_nspath, snp_nsstat) #clean memory

# Filter NS genes not called in dndscv
#skip significance (pmis_cv < 0.05)
#discard nitrogen (not enough genes)
snp_nsgenes <- snp_nsgenes %>%
  dplyr::filter(!(is.na(wmis_cv)) & wmis_cv < 1 & wmis_cv != 0 & pathway != "Nitrogen") %>%
  dplyr::mutate(product = str_replace_all(product, "%2C", ",")) #change %2C to comma

@@

# Merge NS genes to unique selected gene products
snp_nsgenes_sub <- snp_nsgenes %>%
  dplyr::group_by(product, pathway) %>%
  dplyr::summarise(dnds = mean(wmis_cv))
snp_nsgenes_sub_col <- c("#8c2421", "#666666", "#4f4fad", "#ddad3b")

# Lollipop plot of grouped NS gene products
snp_plot_nsgensub <- ggdotchart(snp_nsgenes_sub, x = "product", y = "dnds", color = "pathway",
                                xlab = "NS gene product", ylab = "Mean dN/dS", y.text.col = TRUE,
                                rotate = TRUE, palette = snp_nsgenes_sub_col, dot.size = 3,
                                sorting = "descending", ggtheme = theme_pubr()) +
  scale_y_continuous(limits = c(0, 1)) +
  geom_hline(yintercept = mean(snp_nsgenes$wmis_cv), linetype = 2, linewidth = 0.2) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2)
ggsave("mags-all/5-SNPs/dndscv/results/plot/plot_dnds_nsgenesprod.pdf", snp_plot_nsgensub, height = 9, width = 10)

# Global gene estimation of dN/dS (PS and NS)
snp_estim <- snp_selcv %>%
  dplyr::select(gene_name, wmis_cv, pmis_cv) %>%
  dplyr::filter(wmis_cv != 0)

@@

# Add taxa & QC
snp_estim <- left_join(snp_estim, snp_dnds, by = c("bin")) %>%
  dplyr::select(-c("num", "genus", "species"))

# Normalize dN/dS per gene by MAG contamination, then group by phylum
snp_estim_phy <- snp_estim %>%
  dplyr::mutate(dnds_adj = wmis_cv * (1 - contamination / 100)) %>%
  dplyr::group_by(phylum, bin) %>%
  dplyr::summarise(snps = n(), dnds_sum = sum(dnds_adj)) %>%
  ungroup() %>%
  dplyr::mutate(dnds_glob = dnds_sum / snps) %>% #divide by total genes and get top 3 phyla
  dplyr::filter(phylum == "Proteobacteria" | phylum == "Acidobacteriota" | phylum == "Actinobacteriota")

# Sort phylum by median and set color
snp_estim_sort_phy <- snp_estim_phy %>%
  dplyr::group_by(phylum) %>%
  dplyr::summarise(n = median(dnds_glob)) %>% #by mean?
  dplyr::arrange(desc(n))
snp_estim_sort_phy$color <- rep("#4c6194", nrow(snp_estim_sort_phy))

@@

# Load lists of PS and NS genes (externally* matched against KEGG KOs from KOfamScan hits)
snp_kops <- read.csv("mags-all/5-SNPs/dndscv/results/kofam/ps_genes_kos.txt", sep = "\t", header = FALSE, stringsAsFactors = FALSE)
snp_kons <- read.csv("mags-all/5-SNPs/dndscv/results/kofam/ns_genes_kos.txt", sep = "\t", header = FALSE, stringsAsFactors = FALSE)

# Reduce to unique KOs, saving scores
snp_kops <- snp_kops %>%
  dplyr::group_by(V3, V7) %>%
  dplyr::summarise(sum_ko = n(), score_avg = mean(V4), .groups = "drop") %>%
  dplyr::rename(KO = V3, KO_definition = V7) %>%
  dplyr::mutate(KO_def = paste0("(", KO, ") - ", KO_definition)) %>%
  dplyr::mutate(dNdS = "PS")
snp_kons <- snp_kons %>%
  dplyr::group_by(V3, V7) %>%
  dplyr::summarise(sum_ko = n(), score_avg = mean(V4), .groups = "drop") %>%
  dplyr::rename(KO = V3, KO_definition = V7) %>%
  dplyr::mutate(KO_def = paste0("(", KO, ") - ", KO_definition)) %>%
  dplyr::mutate(dNdS = "NS")

# Merge dataframe and filter by sum
snp_kos <- bind_rows(snp_kops, snp_kons)
snp_kos <- snp_kos %>% dplyr::filter(sum_ko > 10)
rm(snp_kops, snp_kons) #clean memory

@@

# Circos: filter the genome bin with global dN/dS >1
snp_cirmag <- snp_estim %>% dplyr::filter(grepl("S15.49", bin))
